#REDWINE DATA ANALYSIS by Yogeshwar Murugan
========================================================

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
library(dplyr)
library(gridExtra)
library(RColorBrewer)
library(lattice)
library(scales)
library(memisc)
library(reshape)
library(graphics)
library(ggplot2)
library(pander)
```

```{r echo=FALSE, Load_the_Data}
data <- read.csv('wineQualityReds.csv')
str(data)
```

This dataset is public available for research. The details are described in [Cortez et al., 2009].

P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. Modeling wine preferences by data mining from physicochemical properties. In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.

There are 1599 observations and 13 variables in this data set.

Variable description:

Fixed acidity: most acids involved with wine or fixed or nonvolatile (do not evaporate readily)

Volatile acidity: the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste

Citric acid: found in small quantities, citric acid can add 'freshness' and flavor to wines

Residual sugar: the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet

Chlorides: the amount of salt in the wine

Free sulfur dioxide: the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine

Total sulfur dioxide: amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine

Density: the density of water is close to that of water depending on the percent alcohol and sugar content

pH: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale

Sulphates: a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant

Alcohol: the percent alcohol content of the wine

Quality: output variable (based on sensory data, score between 0 and 10)

```{r}
#Creating a new Factored Variable called 'Rating'

data$rating <- ifelse(data$quality < 5, 'bad', ifelse(
  data$quality < 7, 'average', 'good'))

data$rating <- ordered(data$rating,
                       levels = c('bad', 'average', 'good'))
```

# Univariate Plots Section


```{r echo=FALSE, Univariate_Plots}
qplot(x=quality, data=data, geom='bar')
```



Rated on a scale from 1 to 10, has very few exceptionally high or low quality ratings. Qualities 5 and 6 seem to be the highest in count.


```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=pH), data=data)+
   geom_histogram()+
  ggtitle('pH distribution')
```


pH distribution is normaly centered around 3.3 and 3.6


```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=sulphates), data=data)+
   geom_histogram()+
  ggtitle('Sulphates distribution')
```


Sulphates are positively skewed, with highest count ranging between 0.5 to 0.6. There seem to be outliers at 2.0.


```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=chlorides), data=data)+
   geom_histogram()+
  ggtitle('Chlorides distribution')
```

Chlorides are heavily postive skewed, with a center around 0.075.

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=citric.acid), data=data)+
   geom_histogram()+
  ggtitle('Citric Acid')
```

Citric acid looks nearly Trimodal, with a peak at 0.0 and 
nearly constantly decreasing values until another peak value at
0.25 and a third peak at 0.5.

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=total.sulfur.dioxide), data=data)+
   geom_histogram()+
  ggtitle('Total SO2 distribution')
```

Sulfur Dioxide also is heavily positively skewed, with centre at 20 and extending upto 150.

```{r echo=FALSE,warning=FALSE,message=FALSE}
data$total.acidity <- data$fixed.acidity + data$volatile.acidity
ggplot(aes(x=total.acidity), data=data)+
   geom_histogram()+
  ggtitle('Total Acidity (fixed + volatile)')
```


Total acidity of a wine is calcualted by summing up Fixed acidity and Volatile acidity. Hence, 'Total.acidity' was calculated variable by adding 'fixed.acid' and 'volatile.acidty'.  Total Acidity of the wines seem to be slightly positively skewed with maximum number of samples with an acidity around 7.5.


#Univariate Analysis
   
   
###What are the main feature(s) of interest in your dataset?
  I would like to determine which factors determine the quality of a wine. Key varialbes such as 'Volatile Acidity'and 'Alcohol' are the main features of the wine quality.
  
###What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
  'Citric.acid', 'Chlorides' and a few other features might support us in future analysis.

###Did you create any new variables from existing variables in the dataset?
  'Total.Acidity' variable was created from the existing variables as it a key feature.

###Of the features you investigated, were there any unusual distributions? 
  Citric acid was very unusual. It had a trimodal Distribution.
  
###Did you perform any operations on the data to tidy, adjust, or change the form
of the data? If so, why did you do this?
  The dataset was already cleaned and tidy. Only new variables were created.

# Bivariate Plots Section


###Generating a Correlation table for initial insights.

```{r echo=FALSE,warning=FALSE,message=FALSE, Bivariate_Plots}
c <- cor(
  data %>%
    # first we remove unwanted columns
    dplyr::select(-X) %>%
    dplyr::select(-rating) %>%
    mutate(
      # now we translate quality to a number
      quality = as.numeric(quality)
    )
)
emphasize.strong.cells(which(abs(c) > .6 & c != 1, arr.ind = TRUE))
pandoc.table(c)
```

Correlation tables are created for intial insights. 
The following are heavy correlations among the data:

1. pH has a strong negative correlation with fixed acidity.
2. Volatile acidity is higly negatively correlated with the quality of the wine.
3. Density has a very strong correlation with Fixed Acidity.
4. Amount of alcohol considerably correlates with the quality of wine.


```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(data = data, aes(x = pH, y = fixed.acidity)) +
  geom_point( alpha = .3) +
  geom_smooth(alpha = .5,color = 'blue') +
  ggtitle("Fixed acidity vs pH")
```

The feature pH has a strong negative correlation with fixed acidity. 
The truth value of this claim is high as it is known popularly that as 
pH value decreases, acidity increases.

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(data = data, aes(x = quality, y = volatile.acidity)) +
  geom_boxplot() +
  geom_jitter(alpha = 0.3)
  ggtitle("Volatile acidity vs Quality")
```

The feature 'Volatile acidity' is higly negatively correlated with the quality of the wine.
As the Volatile acidity reduces, the quality of the wine seems to increase.

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=density, y=fixed.acidity), data=data) + 
  geom_jitter(alpha=2/3) + 
  geom_smooth() + 
  ggtitle("Density vs Fixed acidity")
```


As Density increases, the fixed acidity of the wine also increases, making sense of a positive correlation.


```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=alcohol, y=quality), data=data) + 
  geom_jitter(alpha=2/3) + 
  geom_smooth() + 
  ggtitle("Alcohol Content vs Quality")
```

Alcohol and quality appear to be slightly correlatable.
Lower quality wines tended to have lower alcohol contents.





```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=alcohol), data=data) + 
  geom_density() + 
  facet_wrap("rating") + 
  ggtitle("Alcohol Content for Wine Quality Ratings") +
  xlab("Alcohol Percentage (%)") +
  ylab("Density (g/dm3) ")
```



Alcohol content definitely affects the wine quality rating.
The effect is not a strong one, and even for a given rating, there is 
significant variability.

# Bivariate Analysis

  Correlation table helped to understand the correlation among different features.
Quality is strongly correlated positively with alcohol. Good wines have lower 
pH values, which also goes with having more fixed acidity.Scatterplot relationships
between these properties often showed a slight trend within the bulk of property values.

  Citric acid and fixed acidity have a strong positive correlation of 0.7, while citrict acid and volatile acidity have a moderate negative correlation of -0.6.
Density and fixed acidity are two features with strong positive correlation of 0.7.
Negative correlation between alcohol and density.
An expected strong negative correlation between pH and fixed and citric acid.
A surprising positive correlation between pH and volatile acidity, since a higher pH value means less acidity, but a higher volatile acidity means more acidity.

The Negative Correlation between pH and fixed acity was the strongest relationship found.

# Multivariate Plots Section


```{r echo=FALSE,warning=FALSE,message=FALSE, Multivariate_Plots}
ggplot(aes(x=alcohol, y=chlorides, color=factor(quality)), data=subset(data,
          data$chlorides < quantile(data$chlorides, 0.99))) + 
  geom_point(position='jitter') +
  facet_wrap(~rating) +
  ggtitle("Alcohol Content vs Chlorides and Wine Quality Ratings")
```

Adding a third varialbe chloride to the Alcohol vs Quality added little reversal of 
insight to the plotUnusually bright chloride points occured at different qualities and alcohol
content points with no discernable pattern. Higher alcohol content and lower
chloride content appears to correlate to higher quality wines.

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(aes(x=citric.acid, y=pH, color=factor(quality)),data=subset(data,
          data$citric.acid < quantile(data$citric.acid, 0.99))) +
  geom_point(position='jitter') +
  geom_smooth() +
  ggtitle("Citric Acid vs pH and Wine Quality Ratings")
```


Higher alcohol content and higher citric acid content appears to correlate to
higher quality wines. pH has no notable effect on wine quality

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(data = data, aes(x = volatile.acidity, y = alcohol)) + 
    facet_wrap(~rating) + 
    geom_point(aes(color = factor(quality)))
```

It looks like Volatile acid has just the negative effect. Good quality redwines
have low concentrations of volatile acid and high amounts of alcohol.

```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(data = data, aes(x = citric.acid, y = volatile.acidity)) +
    facet_wrap(~rating) + 
    geom_point(aes(color = factor(quality)))
```

High amounts of Citric Acid and low Volatile Acids seems to produce better Wines.


```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(data = data,
       aes(y = pH, x = alcohol,
           color = factor(quality))) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating)
```

Low pH and Higher alcohol amounts produce better quality wines.


```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(data = data,
       aes(y = total.sulfur.dioxide, x = alcohol,
           color = factor(quality))) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap(~rating)
```

Lower Sulphur Dioxide seems to produces better wine even though there are some high 
outliers for better wine with high Sulphur Dioxide.



```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(data = data,
       aes(y = sulphates, x = alcohol,
           color = factor(quality))) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap('rating') +
  scale_y_continuous(limits=c(0.3,1.5)) +
  ylab("potassium sulphate (g/dm3)") +
  xlab("Alcohol Percentage (%)") +
  ggtitle("Alcohol and sulphates over wine quality")
```

The best quality wines have high values for both Alcohol percentage and 
Sulphate concentration implying that High alcohol contents and high sulphate
concentrations together seem to produce better wines.


```{r echo=FALSE,warning=FALSE,message=FALSE}
ggplot(data = subset(data, rating != 'average'),
       aes(x = volatile.acidity, y = alcohol, color = rating)) + 
  geom_jitter(size = 2, alpha = 1/2) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  labs(x = "Volatile Acidity (g/dm3)",
       y = "Alcohol (%)",
       title = "Alcohol vs. Volatile Acidity on Wine Quality")
```

High volatile acidity made wine quality down. 
We notice that the lines make is more clear to see the relationship alcohol
and volatile acidity by rating.


# Multivariate Analysis


  High alcohol contribute to good quality of wine, adding citric acid will influence the quality of wine positively, while adding volatile acid influences the quality of wine negatively.Typically, pH is considered important when assessing wine quality, however the data does not show an appreciable correlation


 Citric Acid, even though weakly correlated plays a part in improving the wine quality.

------

# Final Plots and Summary


### Plot One
```{r echo=FALSE,warning=FALSE,message=FALSE, Plot_One}
ggplot(aes(x=alcohol), data=data) + 
  geom_density() + 
  facet_wrap("rating") + 
  ggtitle("Alcohol Content for Wine Quality Ratings") +
  xlab("Alcohol Percentage (%)") +
  ylab("Density (g/dm3) ")
```



### Description One
Alcohol content definitely affects the wine quality rating.
The effect is not a strong one, and even for a given rating, there is 
significant variability. The plot clearly visualizes the shift in alcohol 
percentage as the quailty of the wine gets better.

### Plot Two
```{r echo=FALSE,warning=FALSE,message=FALSE, Plot_Two}
ggplot(data = data,
       aes(y = sulphates, x = alcohol,
           color = factor(quality))) +
  geom_point(alpha = 0.8, size = 1) +
  geom_smooth(method = "lm", se = FALSE,size=1)  +
  facet_wrap('rating') +
  scale_y_continuous(limits=c(0.3,1.5)) +
  ylab("potassium sulphate (g/dm3)") +
  xlab("Alcohol Percentage (%)") +
  ggtitle("Alcohol and sulphates over wine quality")
```

### Description Two
 The best quality wines have high values for both Alcohol percentage and 
Sulphate concentration implying that High alcohol contents and high sulphate
concentrations together seem to produce better wines. The Good quality wine seems to
be slightly downwards, since it the sulphate level tends to be very slighty low as quality increases.

### Plot Three
```{r echo=FALSE,warning=FALSE,message=FALSE, Plot_Three}
ggplot(data = subset(data, rating != 'average'),
       aes(x = volatile.acidity, y = alcohol, color = rating)) + 
  geom_jitter(size = 2, alpha = 1/2) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, size = 1) +
  labs(x = "Volatile Acidity (g/dm3)",
       y = "Alcohol (%)",
       title = "Alcohol vs. Volatile Acidity on Wine Quality")
```

### Description Three
We had to subset the data to take off the 'average' wines as it was just 
overplotting the chart.High volatile acidity made wine quality down. 
We notice that the lines make is more clear to see the relationship alcohol
and volatile acidity by rating. High quality wine have a combination of high 
alcohol content and low volatile acidity.
------

# Reflection

In this dataset more than 70% of the wines are of avergae quality betwen 5 & 6. 
If we could have a dataset of more observations and uniform quality of wines. 
Then we will be able to perform a better analysis. In the final part of the analysis, 
multivariate plots were plotted to see if there were some interesting combinations
of variables which together affected the overall quality of the wine. For future analysis,
better features of the wine samples such as the tastes of the wines, could be given which 
could lead us to a better analysis between the alcoholic and sugar content of the wines.